The promise of community-driven preprints in ecology and evolution

### 2024-06-03 ms/ms.qmd
Up to Breaking down barriers

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has grown to become a popular preprint server for researchers in ecology and evolutionary biology. The server has since been adopted by the California Digital Library (CDL) to be hosted in perpetuity without fees. Editors are researchers from across the globe (currently eight different countries) in the fields of ecology and evolutionary biology who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to post both preprints and postprints, or author-accepted manuscripts -  a version of an accepted or published article without typesetting and formatting by a journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a policy that recognises pre- and post-prints -- though most do; see https://www.sherpa.ac.uk/romeo/).

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published monthly to *EcoEvoRxiv* between 2018 and 2023. *EcoEvoRxiv* was established in June 2018 before the soft launch in November 2018. Notable milestones include when *EcoEvoRxiv* transitioned to the California Digital Library (CDL), the acceptance of Spanish and Portuguese preprints and postprints and when the first IUCN Red list Ecosystem report was accepted; B) Countries that are posting articles (preprints and postprints) to *EcoEvoRxiv* as inferred from the country of affiliation of the submitting author; C) Taxa used / covered in the articles posted to *EcoEvoRxiv* (*n* = 1080 papers); D) Types of preprints accepted on *EcoEvoRxiv* (*n* = 620 papers). E) Career stage of authors posting to *EcoEvoRxiv* along with the gender of the submitting author.
#| echo: false
#| warning: false
#| message: false

@@

To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints as 'articles' in our overview analysis. After removing five duplicate titles – suggesting a few authors created multiple submissions for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55--60 preprints published bi-monthly in the last two years ([@fig-summary]A).  *EcoEvoRxiv* hosts papers from authors based in `r length(unique(countries$country))-1` countries, with 90% of the articles coming from just `r nrow(most_use)-1` countries. North America, Australia, and European countries upload the most preprints with much fewer coming from countries in Africa, Central America, and parts of Asia ([@fig-summary]B).  Articles covered all major taxonomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%), and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Given that *EcoEvoRxiv* is a community-driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses or for practitioners.  As a result, articles on *EcoEvoRxiv* are more diverse than those on preprint servers that have more restrictive submission policies; nonetheless, typical research articles are still the most common type of preprint (`r sum_types[sum_types$type == "research article",]$perc`%), followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]D).

Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem whereby data that are relevant for research syntheses are not published in typical peer-reviewed journals[@haddaway2015shades; @haddaway2020eight]. This is reflected by `r nt`% of articles on *EcoEvoRxiv* that could be considered ‘grey literature’ in ecology and evolutionary biology: books, book chapters, reports, and other research output types. Currently, *EcoEvoRxiv* does not host many reports (`r sum_types[sum_types$type == "report", "perc"]`%), particularly from governments or industry. We would encourage authors to consider posting books, book chapters and reports to ensure they are openly accessible and more easily found. *EcoEvoRxiv* articles are given a unique DOI and indexed on Google Scholar. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) as a place where IUCN reports can be hosted to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these documents meet the IUCN requirements.

>>>>

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary Biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has become a popular preprint server for ecologist and evolutionary biologists. The server has since been adopted by the California Digital Library (CDL) to be hosted in perpetuity without fees. Editors are ecologists and evolutionary biologists from across the globe who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to post both preprints and postprints (also known as author-accepted manuscripts). While preprints are versions of manuscripts by authors before peer-review, postprints are versions of peer-reviewed and accepted articles but without typesetting and formatting by a journal. The main reason for publishing postprints on a preprint server is to ensure published articles are openly accessible to everyone without a paywall (i.e., green open access). The publication of postprints can be done anytime after publication, provided that journals allow it (which most do; see https://www.sherpa.ac.uk/romeo/).

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published on EcoEvoRxiv between 2018 and 2023. EcoEvoRxiv was established in June 2018 before the launch in November 2018. Notable milestones include EcoEvoRxiv transitioning to the California Digital Library (CDL), the acceptance of preprints and postprints in Spanish and Portuguese, and the acceptance of the first IUCN Red List Ecosystem report; B) Geographic origin of the preprints and postprints uploaded to EcoEvoRxiv, inferred from the country of affiliation of the submitting author; C) Taxa used/covered in the articles posted to EcoEvoRxiv (n = 1080 articles); D) Types of preprints accepted on EcoEvoRxiv (n = 620 articles). E) Career stage of authors posting to EcoEvoRxiv along with the gender of the submitting author. Values lower than zero are indicative of authors who uploaded preprints before their first scientific publication in a journal.

#| echo: false
#| warning: false
#| message: false

@@

To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). We consider both preprints and postprints as ‘articles’. After removing five duplicate titles – suggesting that a few authors created multiple submissions for the same preprint rather than updating the existing submission – we had data for a total of `r pr` articles with ~55--60 preprints published bi-monthly in the last two years ([@fig-summary]A).  *EcoEvoRxiv* hosts articles from authors based in `r length(unique(countries$country))-1` countries, with 90% of the articles coming from just `r nrow(most_use)-1` countries. North America, Australia, and European countries upload the most preprints with many fewer coming from countries in Africa, Central America, and parts of Asia ([@fig-summary]B).  Articles covered all major taxonomic groups, with the most common being vertebrates (`r round(sum_taxa$perc[6], digits = 1)`%), plants (`r round(sum_taxa$perc[1], digits = 1)`%), and invertebrates (`r round(sum_taxa$perc[5], digits = 1)`%) ([@fig-summary]C).

@@

Given that *EcoEvoRxiv* is a community-driven preprint server it has managed to diversify the types of articles accepted. *EcoEvoRxiv* accepts reports and books that are often not easily accessible and may contain important data relevant to research synthesists or practitioners. As a result, articles on *EcoEvoRxiv* are more diverse than those on preprint servers that have more restrictive submission policies. For example, *bioRxiv* only accepts research articles (https://www.biorxiv.org/submit-a-manuscript). Typical research articles are still still the most common type of preprint (`r round(sum_types[sum_types$type == "research article",]$perc, digits = 1)`%), followed by reviews and meta-analyses (`r round(sum_types[sum_types$type == "reviews and meta-analyses",]$perc, digits = 1)`%) and opinion papers (`r round(sum_types[sum_types$type == "opinion",]$perc, digits = 1)`%) ([@fig-summary]D).

Accepting a greater diversity of article types allows *EcoEvoRxiv* to help deal with the 'grey literature' problem, whereby data that are relevant for research syntheses are not published in typical peer-reviewed journals[@haddaway2015shades; @haddaway2020eight]. This is reflected by `r nt`% of the articles on *EcoEvoRxiv* being books, book chapters, reports, and other research output types, which are typically considered ‘grey literature’ in ecology and evolutionary biology. Currently, *EcoEvoRxiv* does not host many reports, particularly from government or industry. We would encourage authors to consider posting books, book chapters, and reports to ensure that they are openly accessible and more easily found. *EcoEvoRxiv* articles are given a unique DOI and indexed on Google Scholar. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) as a place where IUCN reports can be hosted to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these documents meet the IUCN requirements.
